---
title: "Wordle DataAnalysis Report"
author: "Brenda Luis"
date: "2025-12-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Introduction 

Wordle is a daily word puzzle in which players attempt to guess a five-letter English word within six attempts. Each guess provides structured feedback based on letter correctness and position. Although the game appears simple, its dataset of solutions contains linguistic patterns that can be statistically analyzed.

The goal of this project is to understand the structure of the official Wordle answer list. By examining letter usage, positional behavior, bigram patterns, repeated-letter distributions, and relationships between word frequency and Scrabble score, this project answers the broader question:

**Does Wordle follow predictable linguistic patterns, and how can these patterns improve player strategy?**

This problem is meaningful because understanding these trends allows players to make more informed guesses and reveals how puzzle designers curate word lists.

# Questions to Answer
This analysis focuses on the following core questions: 

1. Which letters appear more frequently in Wordle solutions?
2. Which letters appear more frequently in Wordle solutions by position?
3. Which letter pairs (bigrams) appear most often in Wordle solutions?
4. How does the bigram frequency change depending on where they appear in the word?
5. Do repeated letter Wordle answers follow seasonal or monthly patterns across different years?
6. Is there a relationship between a word’s frequency in English and its Scrabble point value?



# Data Preparation


```{r message=FALSE, warning=FALSE}
# libraries 
library(ggplot2)    # plot visualization
library(dplyr)      # summarizing, filtering, etc
library(tidyr)      # reshaping data (pivot, etc)
library(kableExtra) # table formatting
library(gridExtra)  # arranging plots
library(patchwork)  #
library(calendR)    # generating monthly heatmaps
library(lubridate)
library(ggthemes)
```

# Data Analysis


### Question 1
question background and info 

```{r}
# code
```
interpretation of output. 




### Question 2
question background and info 

```{r}
# code
```
interpretation of output. 



# Summary 


