---
title: "Wordle Data Analysis"
author: "Brenda Luis"
date: "2025-11-09"
output: slidy_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Introduction

Wordle is a popular online word puzzle where players guess a five letter word within six attempts. Each guess provides feedback: letters in the correct position are highlighted in green, letters in the word but in the wrong position appear yellow, and letters not in the word are gray. Beyond being an engaging game, Wordle presents a rich dataset for analysis. Words have unique characteristics, such as vowel-consonant balance, repeated letters, rare letters, palindromes, Scrabble scores, and frequency of usage.

This analysis explores these patterns across 1600+ Wordle words to uncover trends in word selection, letter placement, and other structural features. By examining the relationships between word properties, frequency, and temporal trends, this project aims to provide insights into the hidden patterns that shape the daily puzzle.

## Questions to Answer

-

## Data Preparation

```{r cars, echo = TRUE}
library(ggplot2) # plot visualization
library(dplyr)   # data analysis (summarizing, filtering, etc)
library(tidyr)   # reshaping data (pivot, etc)
library(stringr) # string manipulation (used for bigrams) ok actually i realized paste would work for my bigram graphs but keeping this library in case i need to do other string stuff 
```

```{r}
# I scrapped this data from 
# https://wordfinder.yourdictionary.com/wordle/answers/
# and added what I thought to be relevant data columns 

# some column names are easy to understand 
# I will add some section explaining what the dataset columns mean

wordle_data <- read.csv("wordle_data.csv")
glimpse(wordle_data)
```
## Letter Counts

letter counts overall
show trends on most used letters


note: my own scores. how many tries did it take me 


```{r}
wordle_data


letter_frequency <- wordle_data %>% 
  select(P1, P2, P3, P4, P5) %>% 
  pivot_longer(cols = c(P1, P2, P3, P4, P5), names_to = "Position", values_to = "Letter") %>% 
  group_by(Letter) %>% 
  summarise(Counts = n()) %>% 
  arrange(-Counts)
  

head(letter_frequency, n = 4)
tail(letter_frequency, n = 4)

letter_frequency

letter_frequency %>% 
  ggplot(aes(x = reorder(Letter,-Counts), y = Counts)) + geom_col()
```


## Letter Counts by position

letter counts by positions
show trends on most used letters by position 

The heatmap shows how frequently each letter appears in each position of Wordle answers. Darker shades indicate more common letters. By visualizing letter frequency across positions we can identify common structural patterns in five letter words like certain consonants often starting words.

S appears very often in P1 (starting position)
Y is very common in P5



```{r}
# x = letter
# y = position 
# heatmap 


positional_letter_frequency <- wordle_data %>% 
  select(P1, P2, P3, P4, P5) %>% 
  pivot_longer(cols = c(P1, P2, P3, P4, P5), names_to = "Position", values_to = "Letter") %>% 
  group_by(Position, Letter) %>% 
  summarise(Counts = n()) %>% 
  arrange(Position, -Counts)

positional_letter_frequency

positional_letter_frequency %>% 
  ggplot(aes(x = reorder(Letter, -Counts), y = Counts)) + 
  geom_col() + 
  facet_wrap(~Position, scales = "free_x")

positional_letter_frequency %>% 
  ggplot(aes(x = reorder(Letter, -Counts), y = Counts)) + 
  geom_col() + 
  facet_wrap(~Position)

positional_letter_frequency %>%
  ggplot(aes(x = Letter, y = Position, fill = Counts)) +
  geom_tile(color = "White") + 
  geom_text(aes(label = Counts), color = "White", size = 2) +
  scale_fill_gradient(low = "salmon", high = "darkred")

# filter and plot
# vowels first 
# or maybes just vowels

```





## Data Analysis

```{r pressure}
# so I found this data on kaggle
# https://www.kaggle.com/code/bcruise/wordle-letter-frequencies
# and towards the end they do common bigram analysis 
# will do that with my data 
# so I can use P1, P2, P3, P4, P5
# this holds the individual letters by position

bigrams <- wordle_data %>% 
  mutate (B1 = paste0(P1, P2),  # paste0 no separator by default
          B2 = paste0(P2, P3),
          B3 = paste0(P3, P4),
          B4 = paste0(P4, P5)) %>% 
  select(B1, B2, B3, B4) %>% 
  pivot_longer(cols = everything(), names_to = "Position", values_to = "Bigram") %>% # names_to = col name to create from specified cols before, values_to = new column to create from values in each cell.
  group_by(Bigram) %>% 
  summarise(Counts = n()) %>% 
  arrange(-Counts)
 


bigrams

```

## MOST COMMON BIGRAMS 
position not taken  into account here

what does this graph tell us? 
the Most Common Bigrams plot shows which two letter combos appear most frequently across all Wordle answers regardless of where they occur in the word. This gives insight into the most common letter pairings in English five letter words which can be useful for players choosing strong starting guesses.

```{r}

# plotting bigram data
# code below bad 
# didnt realize i had 354 possible bigrams

# bigrams %>% ggplot(aes(x = Bigram)) + geom_bar()

# instead lets take top bigrams 
top_bigrams <- bigrams %>% head(25)
top_bigrams

# top_bigrams %>% ggplot(aes(x = Bigram)) + geom_bar()

# https://r-resources.massey.ac.nz/rgcookbook/RECIPE-BAR-GRAPH-COUNTS.html
# section 3.3.4

# geom_col uses stat_identity by default so I can pass in the pre-calculated counts
top_bigrams %>% ggplot(aes(x = reorder(Bigram, -Counts), y = Counts)) + geom_col()

# will add labels and stuff later. 
```

## Most common bigrams by position?

The Most Common Bigrams by Position plot breaks bigrams data down by where the pair appears within the word. This helps reveal structural patterns like certain bigrams like “TH” or “SH” can appear more often at the start of words while others like “ER” or “LY” can appear near the end more frequently. Understanding these trends can guide more strategic guesses by considering both frequency and position. I will go deeper into this when looking at the prefix and suffix columns in the data

```{r}

bigrams_by_position <- wordle_data %>% 
  mutate (B1 = paste0(P1, P2),  # paste0 no separator by default
          B2 = paste0(P2, P3),
          B3 = paste0(P3, P4),
          B4 = paste0(P4, P5)) %>% 
  select(B1, B2, B3, B4) %>% 
  pivot_longer(cols = everything(), names_to = "Position", values_to = "Bigram") %>%  # names_to = col name to create from specified cols before, values_to = new column to create from values in each cell.
  group_by(Position, Bigram) %>% 
  summarise(Counts = n()) %>% 
  arrange(Position, -Counts)
 
bigrams_by_position
```

```{r}
# just taking 10 from each position
# b4 has a tie

top_bigrams_by_pos <- bigrams_by_position %>%
  group_by(Position) %>%
  slice_max(Counts, n = 10)

top_bigrams_by_pos

# positional bigrams do not share same x axis :(
top_bigrams_by_pos %>% ggplot(aes(x = Bigram, y = Counts)) + geom_col() + facet_wrap(~Position)

# fixed here
# ok why is this not reordering?? 
top_bigrams_by_pos %>% ggplot(aes(x = reorder(Bigram, -Counts), y = Counts)) + geom_col() + facet_wrap(~Position, scales = "free_x")
```


# Summary

