---
title: "Wordle DataAnalysis Report"
author: "Brenda Luis"
date: "2025-12-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Introduction 

Wordle is a popular online word puzzle where players guess a five letter word within six attempts. Each guess provides feedback: letters in the correct position are highlighted in green, letters in the word but in the wrong position appear yellow, and letters not in the word are gray. Beyond being an engaging game, Wordle presents a rich dataset for analysis. Words have unique characteristics, such as vowel-consonant balance, repeated letters, rare letters, Scrabble scores, and frequency of usage.

This analysis explores these patterns across 1600+ Wordle words to uncover trends in word selection, letter placement, and other structural features. By examining the relationships between word properties, frequency, and temporal trends, this project aims to provide insights into the hidden patterns that shape the daily puzzle.

# Questions to Answer
This analysis focuses on the following core questions: 

1. Which letters appear more frequently in Wordle solutions?
2. Which letters appear more frequently in Wordle solutions by position?
3. Which letter pairs (bigrams) appear most often in Wordle solutions?
4. How does the bigram frequency change depending on where they appear in the word?
5. Do repeated letter Wordle answers follow seasonal or monthly patterns across different years?
6. Is there a relationship between a word’s frequency in English and its Scrabble point value?



# Data Preparation


```{r}
# libraries 
# need to hide output 
library(lubridate)
```

# Data Analysis


### Question 1
question background and info 

```{r}
# code
```
interpretation of output. 




### Question 2
question background and info 

```{r}
# code
```
interpretation of output. 



# Summary 


