---
title: "Wordle Data Analysis"
author: "Brenda Luis"
date: "2025-11-09"
output: slidy_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Introduction

Wordle is a popular online word puzzle where players guess a five-letter word within six attempts. Each guess provides feedback: letters in the correct position are highlighted in green, letters in the word but in the wrong position appear yellow, and letters not in the word are gray. Beyond being an engaging game, Wordle presents a rich dataset for analysis. Words have unique characteristics, such as vowel-consonant balance, repeated letters, rare letters, palindromes, Scrabble scores, and frequency of usage.

This analysis explores these patterns across 1600+ Wordle words to uncover trends in word selection, letter placement, and other structural features. By examining the relationships between word properties, frequency, and temporal trends, this project aims to provide insights into the hidden patterns that shape the daily puzzle.

## Questions to Answer

- Bullet 1
- Bullet 2
- Bullet 3

## Data Preparation

```{r cars, echo = TRUE}
library(ggplot2) # plot visualization
library(dplyr)   # data analysis (summarizing, filtering, etc)
library(tidyr)   # reshaping data (pivot, etc)
library(stringr) # string manipulation (used for bigrams)
```

```{r}
# I scrapped this data from 
# https://wordfinder.yourdictionary.com/wordle/answers/
# and added what I thought to be relevant data columns 

# some column names are easy to understand 
# I will add some section explaining what the dataset columns mean

wordle_data <- read.csv("wordle_data.csv")
glimpse(wordle_data)
```


## Data Analysis

```{r pressure}
# so I found this data on kaggle
# https://www.kaggle.com/code/bcruise/wordle-letter-frequencies
# and towards the end they do common bigram analysis 
# will do that with my data 
# so I can use P1, P2, P3, P4, P5
# this holds the individual letters by position

bigrams <- wordle_data %>% 
  mutate (B1 = paste0(P1, P2),  # paste0 no separator by default
          B2 = paste0(P2, P3),
          B3 = paste0(P3, P4),
          B4 = paste0(P4, P5)) %>% 
  select(B1, B2, B3, B4) %>% 
  pivot_longer(cols = everything(), names_to = "Position", values_to = "Bigram") %>% # names_to = col name to create from specified cols before, values_to = new column to create from values in each cell.
  group_by(Bigram) %>% 
  summarise(Counts = n()) %>% 
  arrange(-Counts)
 


bigrams

```


```{r}
# plotting bigram data
# code below bad 
# didnt realize i had 354 possible bigrams

bigrams %>% ggplot(aes(x = Bigram)) + geom_bar()

# instead lets take top bigrams 
top_bigrams <- bigrams %>% head(25)
top_bigrams

top_bigrams %>% ggplot(aes(x = Bigram)) + geom_bar()

# https://r-resources.massey.ac.nz/rgcookbook/RECIPE-BAR-GRAPH-COUNTS.html
# section 3.3.4

# geom_col uses stat_identity by default so I can pass in the pre-calculated counts
top_bigrams %>% ggplot(aes(x = reorder(Bigram, -Counts), y = Counts)) + geom_col()

# will add labels and stuff later. 
```


# Summary

