---
title: "Wordle DataAnalysis Report"
author: "Brenda Luis"
date: "2025-12-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Introduction 

Wordle is a daily word puzzle in which players attempt to guess a five-letter English word within six attempts. Each guess provides structured feedback based on letter correctness and position. Although the game appears simple, its dataset of solutions contains linguistic patterns that can be statistically analyzed.

The goal of this project is to understand the structure of the official Wordle answer list. By examining letter usage, positional behavior, bigram patterns, repeated-letter distributions, and relationships between word frequency and Scrabble score, this project answers the broader question:

**Does Wordle follow predictable linguistic patterns, and how can these patterns improve player strategy?**

This problem is meaningful because understanding these trends allows players to make more informed guesses and reveals how puzzle designers curate word lists.

<hr style="border:1px solid gray">

# Questions to Answer
This analysis focuses on the following core questions: 

1. Which letters appear more frequently in Wordle solutions?
2. Which letters appear more frequently in Wordle solutions by position?
3. Which letter pairs (bigrams) appear most often in Wordle solutions?
4. How does the bigram frequency change depending on where they appear in the word?
5. Do repeated letter Wordle answers follow seasonal or monthly patterns across different years?
6. Is there a relationship between a word’s frequency in English and its Scrabble point value?

<hr style="border:1px solid gray">

# Required Packages

```{r message=FALSE, warning=FALSE}
# libraries 
library(ggplot2)    # plot visualization
library(dplyr)      # summarizing, filtering, etc
library(tidyr)      # reshaping data (pivot, etc)
library(kableExtra) # nice looking tables 
library(gridExtra)  # arranging multiple plots next to each other
library(patchwork)  # another tool to combine plots
library(calendR)    # generating monthly heatmaps
library(lubridate)  # date handling
library(ggthemes)   # different plot themes
```

<hr style="border:1px solid gray">

# Dataset Information {.tabset}
The dataset contains official Wordle answers from 2021 to 2025, scraped from:

<https://wordfinder.yourdictionary.com/wordle/answers/>

Each row represents one day’s solution.

wordle_data is a data frame with 1609 rows and 28 variables:


```{r echo=FALSE}
variable_info <- data.frame(
  Variable = c(
    "Date","Wordle.Number","Answer","Weekday","Month","Day","Year",
    "Has.Repeated.Letters","All.unique.letters","Starts.with.vowel",
    "Ends.with.vowel","Num.vowels","Num.consonants","Ends.with.Y",
    "Has.rare.letter","Contains.common.prefix","Contains.common.suffix",
    "P1–P5","Word.Frequency","Scrabble.score","Frequency.category",
    "Top.row.letters","Mid.row.letters","Bottom.row.letters"
  ),
  Type = c(
    "date","integer","character","character","character","integer","integer",
    "integer","integer","integer","integer","integer","integer","integer",
    "integer","integer","integer","character","numeric","integer",
    "character","integer","integer","integer"
  ),
  Description = c(
    "Publication date of the puzzle",
    "Sequential ID of the puzzle",
    "Correct 5-letter Wordle answer",
    "Day of the week",
    "Month name",
    "Day of month",
    "Year",
    "1 if the word contains repeated letters",
    "1 if all letters in the word are unique",
    "1 if the word starts with a vowel",
    "1 if the word ends with a vowel",
    "Number of vowels",
    "Number of consonants",
    "1 if the word ends with 'Y'",
    "1 if the word contains Q, Z, X, or J",
    "1 if it contains RE, UN, IN, IM, DE, EX, or BE",
    "1 if it contains ED, ER, LY, ES, ST, NT, TH, CH, SH",
    "Letters in each of the 5 positions",
    "English frequency score (0–1)",
    "Scrabble point value",
    "Frequency category",
    "Count of top-row keyboard letters",
    "Count of middle-row keyboard letters",
    "Count of bottom-row keyboard letters"
  )
)

```

```{r echo=FALSE}
variable_info %>% 
  kbl() %>% 
  kable_styling(full_width = TRUE) %>% 
  scroll_box(width = "100%", height = "350px")

```

<hr style="border:1px solid gray">

# Data Loading

```{r}
# load data from csv
wordle_data <- read.csv("wordle_data.csv")

# convert Date column from string to date object
wordle_data <- wordle_data %>% 
  mutate(Date = as.Date(Date, format = "%Y-%m-%d"))

# convert Month and Weekday columns to factors
wordle_data <- wordle_data %>%
  mutate(
    Month = factor(Month, levels = month.name, ordered = TRUE),
    Weekday = factor(Weekday, levels = c("Sunday","Monday","Tuesday","Wednesday","Thursday","Friday","Saturday"), ordered = TRUE)
  )

# display random sample from data
wordle_data %>%  
  sample_n(10) %>% 
  arrange(Date) %>% 
  kbl() %>% 
  kable_styling() %>% 
  scroll_box(width = "100%", height = "400px")
```


<hr style="border:1px solid gray">

# Data Analysis 


### Question 1 - Which letters appear more frequently in Wordle solutions? {.tabset}
This project begins by looking at how often each letter appears across all Wordle solutions. Even though every answer is a five-letter English word, some letters show up much more often than others. Checking this overall distribution helps us understand the general patterns Wordle follows. Common letters make certain guesses stronger, while rare letters can make a puzzle harder if they appear. The table and bar chart below show how each letter is used in the full list of solutions.

#### letter frequencies table
```{r message=FALSE, warning=FALSE}
# pulling letters, counts, and type
letter_frequency <- wordle_data %>% 
  select(P1, P2, P3, P4, P5) %>% 
  pivot_longer(cols = c(P1, P2, P3, P4, P5), names_to = "Position", values_to = "Letter") %>% 
  group_by(Letter) %>% 
  summarise(Counts = n()) %>% 
  arrange(-Counts) %>% 
  mutate(Type = ifelse(Letter %in% c("A","E","I","O","U"), "Vowel", "Consonant"))

# table displaying data
letter_frequency %>%  
  kbl() %>% 
  kable_styling() %>% 
  scroll_box(width = "40%", height = "400px")
```

#### frequencies bar graph 
```{r}
# bar graph to display letter frequency
letter_frequency %>% 
  ggplot(aes(x = reorder(Letter,-Counts), y = Counts, fill = Type)) + 
  geom_col() +
  labs(
    title = "Frequency of Letters in Wordle Solutions",
    x = "Letter",
    y = "Count",
    fill = "Type"
  ) + 
  scale_fill_manual(values = c("Vowel" = "#485c63", "Consonant" = "#8facb5")) + 
  theme_tufte() + 
  theme(
    plot.title = element_text(size = 20, face = "bold"),   # title size
    axis.title.x = element_text(size = 20),                # X label
    axis.title.y = element_text(size = 20),                # Y label
    axis.text.x = element_text(size = 15), # letter labels
    axis.text.y = element_text(size = 15),                  # position numbers
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 12)
  )
  
```

#### most common letter
```{r}
# most common letter
letter_frequency %>% 
  summarise(most_common_letter = Letter[which.max(Counts)],
            frequency = max(Counts))%>% 
  kbl() %>% 
  kable_styling() %>% 
  scroll_box(width = "50%")

letter_frequency %>% 
  summarise(least_common_letter = Letter[which.min(Counts)],
            frequency= min(Counts))%>% 
  kbl() %>% 
  kable_styling() %>% 
  scroll_box(width = "50%")
```

#### result interpretation
The results show that E is the most common letter, followed by A, R, O, and T. These letters appear throughout many standard English words, so it makes sense that Wordle uses them often. On the other end, letters like X, Z, Q, and J appear very rarely. Because of this gap between common and uncommon letters, first guesses that include frequent vowels and consonants are usually more effective. This gives us a clearer picture of Wordle’s overall word patterns before we look at *where* the letters appear inside the words.




### Question 2
question background and info 

```{r}
# code
```
interpretation of output. 



# Summary 


